---
title: "64060_Assignment_1"
author: "Bryan Mack"
date: "2025-06-01"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.  Download a dataset from the web. You may use any source, but specify the source in your code. Also ensure that the data has a mix of quantitative and qualitative (categorical) variables.
2.  Import the dataset into R (when working in console or RStudio)
    1.  Later imported into RMD Environment (changes "Data" to "Cereals")

```{r}
Cereals <- read.csv("D:/R_DATA/Cereals.csv")
View(Cereals)
```

3.  Print out descriptive statistics for a selection of quantitative and categorical variables.

    ```{r}
    summary(Cereals$protein)
    ```

    ```{r}
    table(Cereals$mfr)
    ```

    ```{r}

    ```

4.  Transform at least one variable. It doesn't matter what the transformation is.

    ```{r echo=TRUE}
    Cereals$dubl_calories <- Cereals$calories * 2
    Cereals$dubl_protein <- Cereals$protein * 2
    Cereals$dubl_fat <- Cereals$fat * 2
    Cereals$dubl_sodium <- Cereals$sodium * 2
    Cereals$dubl_fiber <- Cereals$fiber * 2
    Cereals$dubl_carbo <- Cereals$carbo * 2
    Cereals$dubl_sugars <- Cereals$sugars * 2
    Cereals$dubl_potass <- Cereals$potass * 2
    Cereals$dubl_vitamins <- Cereals$vitamins * 2
    Cereals$dubl_cups <- Cereals$cups * 2
    ```

<!-- -->

5.  Plot at least one quantitative variable, and one scatterplot

```{r}
plot(Cereals$protein,Cereals$calories)
```

```{r}
#load ggplot2 library
library(ggplot2)
ggplot(Cereals, aes(x = calories, y = protein)) +
  geom_point(size = 3, color = "blue") +
  geom_point(aes(x = dubl_calories, y = dubl_protein), size = 3, color = "red", shape = 17) +
  labs(title = "Scatterplot of Calories vs Protein and Their Doubled Values",
       x = "Calories (Blue) / Double Calories (Red)",
       y = "Protein (Blue) / Double Protein (Red)") +
  theme_minimal()
```

6.  Upload your R program, and any associated data files to your git account. Remember to create a separate repository for this class.

    <https://github.com/bryan-mack/64060_bmack>

    <https://github.com/bryan-mack/64060_bmack/blob/main/Assignment_1/Cereals.csv>

7.  Paste the address to your repository in the assignment submission box here in Canvas.

<https://github.com/bryan-mack/64060_bmack/tree/d18b038a2ee0c9309b004d8dd5b0645e550eaefe/Assignment_1>
