---
title: "Assignment_4v1a"
author: "Bryan Mack"
date: "2025-06-14"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dir.create("images")
```

Purpose:
The purpose of this assignment is to use k-Means for clustering.

Directions:
An equities analyst is studying the pharmaceutical industry and would like your help in exploring and understanding the financial data collected by her firm. Her main objective is to understand the structure of the pharmaceutical industry using some basic financial measures. Financial data gathered on 21 firms in the pharmaceutical industry are available in the file Pharmaceuticals.csv. For each firm, the following variables are recorded:

1. Market capitalization (in billions of dollars) 
2. Beta 
3. Price/earnings ratio 
4. Return on equity 
5. Return on assets 
6. Asset turnover 
7. Leverage 
8. Estimated revenue growth 
9. Net profit margin 
10. Median recommendation (across major brokerages) 
11. Location of firm’s headquarters 
12. Stock exchange on which the firm is listed

```{r Library Install}
library(tidyverse)  # data manipulation
#install.packages("factoextra") # if necessary
library(factoextra) # clustering algorithms & visualization
library(ISLR)
library(flexclust)
set.seed(123)
```

```{r Working Directory}
#setwd("D:/R_DATA/Assignment 4")
getwd()
#setwd() #used to set R working directory 
# i.e. D:/R_DATA or (note / for windows OS) 
# setwd("D:\\\\R_DATA\\\\")
```


```{r Data Pharmaceuticals.csv}
#df<-Auto[,c(1,6)]
#summary(df)
#df<-Pharmaceuticals.csv
#df <- read.csv("Pharmaceuticals.csv", stringsAsFactors = FALSE)
getwd()
df <- read_csv("Pharmaceuticals.csv", col_select = c(Market_Cap, Beta, PE_Ratio, ROE, ROA,Asset_Turnover, Leverage, Rev_Growth, Net_Profit_Margin))
#df<-Pharmaceuticals.csv
summary(df)
```

```{r Scale Data}
df <- scale(df, center = TRUE, scale = TRUE)
distance <- get_dist(df)
fviz_dist(distance)
```

```{r}
k4 <- kmeans(df, centers = 4, nstart = 25) # k = 4, number of restarts = 25

# Visualize the output

k4$centers # output the centers

k4$size # Number of cars in each cluster

k4$cluster[10] # Identify the cluster of the 120th observation as an example

fviz_cluster(k4, data = df) # Visualize the output
```

```{r}
#install.packages("flexclust")
#library(flexclust)
set.seed(123)
#kmeans clustering, using manhattan distance
k4 = kcca(df, k=4, kccaFamily("kmedians"))
k4
```

```{r}
#Apply the predict() function
clusters_index <- predict(k4)
dist(k4@centers)
image(k4)
points(df, col=clusters_index, pch=19, cex=0.3)
```

```{r}
library(tidyverse)  # data manipulation
library(factoextra) # clustering & visualization
library(ISLR)
set.seed(123)

df <- scale(df)
fviz_nbclust(df, kmeans, method = "wss")
```

```{r}
fviz_nbclust(df, kmeans, method = "silhouette")
```

```{r}
fviz_nbclust(df, kmeans, method = "gap_stat")
```

Cluster in K5

```{r}
set.seed(123)
#kmeans clustering, using manhattan distance
k5 = kcca(df, k=5, kccaFamily("kmedians"))
k5
```


```{r}
k5 <- kmeans(df, centers = 5, nstart = 25) # k = 4, number of restarts = 25

# Visualize the output

k5$centers # output the centers

k5$size # Number of cars in each cluster

k5$cluster[5] # Identify the cluster of the 120th observation as an example

fviz_cluster(k5, data = df) # Visualize the output
```

```{r}
set.seed(123)
#kmeans clustering, using manhattan distance
k3 = kcca(df, k=3, kccaFamily("kmedians"))
k3
```

```{r}
k3 <- kmeans(df, centers = 3, nstart = 25) # k = 4, number of restarts = 25

# Visualize the output

k3$centers # output the centers

k3$size # Number of cars in each cluster

k3$cluster[3] # Identify the cluster of the 120th observation as an example

fviz_cluster(k3, data = df) # Visualize the output
```

```{r}
k3$centers # output the centers

k3$betweenss # Number of cars in each cluster

k3$cluster[120] # Identify the cluster of the 120th observation as an example

fviz_cluster(k3, data = df) # Visualize the output
```



Tried to change weights of numeric variables. 
```{r}
df1 <- read_csv("Pharmaceuticals.csv", col_select = c(Market_Cap, Beta, PE_Ratio, ROE, ROA,Asset_Turnover, Leverage, Rev_Growth, Net_Profit_Margin))
```
```{r}
df1_scaled <- scale(df1, center = TRUE, scale = TRUE)
df1_scaled <- as.data.frame(scale(df1))
```


```{r}
weights <- c(
  Market_Cap = 2.0,        # Emphasize company size
  Beta = 0.5,              # Lower weight for risk
  PE_Ratio = 0.5,          # Lower weight for valuation
  ROE = 2.0,               # Emphasize profitability
  ROA = 1.0,               # Moderate weight for efficiency
  Asset_Turnover = 0.5,    # Lower weight for operational metric
  Leverage = 0.5,          # Lower weight for debt
  Rev_Growth = 1.0,        # Moderate weight for growth
  Net_Profit_Margin = 2.0  # Emphasize profitability
)
```


```{r}
df1_weighted <- df1_scaled
for (col in names(weights)) {
  df1_weighted[[col]] <- df1_scaled[[col]] * weights[col]
}
```


```{r}
fviz_nbclust(df1_weighted, kmeans, method = "wss")
```


```{r}
fviz_nbclust(df1_weighted, kmeans, method = "silhouette")
```


```{r}
fviz_nbclust(df1_weighted, kmeans, method = "gap_stat")
```

```{r}
set.seed(123)
#kmeans clustering, using manhattan distance
k2_weighted = kcca(df1_weighted, k=2, kccaFamily("kmedians"))
k2_weighted
```

```{r}
k2_weighted <- kmeans(df1_weighted, centers = 2, nstart = 25) # k = 2, number of restarts = 25

# Visualize the output

k2_weighted$centers # output the centers

k2_weighted$size # Number of cars in each cluster

k2_weighted$cluster[120] # Identify the cluster of the 120th observation as an example

fviz_cluster(k2_weighted, data = df1_weighted) # Visualize the output
```


```{r}
set.seed(123)
#kmeans clustering, using manhattan distance
k3_weighted = kcca(df1_weighted, k=3, kccaFamily("kmedians"))
k3_weighted
```

```{r}
k3_weighted <- kmeans(df1_weighted, centers = 3, nstart = 25) # k = 4, number of restarts = 25

# Visualize the output

k3_weighted$centers # output the centers

k3_weighted$size # Number of cars in each cluster

k3_weighted$cluster[120] # Identify the cluster of the 120th observation as an example

fviz_cluster(k3_weighted, data = df1_weighted) # Visualize the output
```




Use cluster analysis to explore and analyze the given dataset as follows:

a. Use only the numerical variables (1 to 9) to cluster the 21 firms. Justify the various choices made in conducting the cluster analysis, such as weights for different variables, the specific clustering algorithm(s) used, the number of clusters formed, and so on. 

I tried several different combinations in clustering including modifying the weights of the data. After I modified the weights I was shown in "fviz_nbclust(df1_weighted, kmeans, method = "silhouette")" the optimal number of clusters was 2 however even after I think I weighted the data to increase the size of the market cap (Market_Cap = 2.0,        # Emphasize company size) relative to other factors I still found exceptions in the clustering. Using the weighted 3 cluster I see in the cluster 1 (assuming the number is the row in the data)
11 - Eli Lilly and Company 
17 - Novartis AG
15 - Medicis Pharmaceutical Corporation (market cap only 1.2)
18 - Pfizer Inc


b. Interpret the clusters with respect to the numerical variables used in forming the clusters. 

Based on the output I am getting I do not think that all of the numerical data should be used for the analysis and that a more specific goal should be defined before the variables are selected for analysis. 

c. Is there a pattern in the clusters with respect to the numerical variables (10 to 12)? (those not used in forming the clusters)

I was most interested in Median_Recommendation however I was not able to see a clear relation between the clusters and the non-numeric data (Median_Recommendation, Location, Exchange)

d. Provide an appropriate name for each cluster using any or all of the variables in the dataset.

Using weighting I was trying to keep all of the variables and expected to see the following 3 categories:

Cluster 1: Large, highly profitable firms.
Cluster 2: Mid-sized, growth-oriented firms.
Cluster 3: Smaller or high-risk firms.
